Identifying Software Performance Changes Across Variants and Versions

### 2020-08-12 paper.tex
text ready

		In this work, we combine two perspectives -- variability and time -- {\color{red} into a novel perspective}. We propose an approach to identify configuration-dependent performance changes: {\color{red}\emph{spatio-temporal performance analysis}}. In a nutshell, we iteratively sample pairs of configurations and versions and measure the respective performance that help us update a model of likelihoods for performance changes. Pursuing a search strategy with the goal of measuring selectively and incrementally further pairs, we increase the accuracy of identified change points related to configuration options and interactions.

@@

	{\color{red}We propose a novel approach, \emph{spatio-temporal performance analysis}}. Our main idea is as follows: We address the configuration complexity of this problem by selecting representative sample sets of configurations.
	Next, we uniformly sample a constant number of commits for each configuration and conduct respective performance measurements.
	Based on these measurement data, we assemble a prediction model.
	For each configuration, we estimate the likelihood of each commit being a change point, i.e. that performance of some configurations changes abruptly compared to the previous commit.
	Next, we leverage similarities in the performance histories of configurations that share common options.
	Since we sample the commits for each configuration independently, for each change point, we obtain many estimations using measurements of different commits.
	Overall, this allows us to obtain more accurate estimations of performance-changing commits with tractable effort.
	Based on a mapping from configurations to predicted change point probabilities for each configuration and commit, we derive the options responsible for each particular change point: configuration-dependent change points.
	In summary, we offer the following contributions:
	\begin{itemize}
		\item A {\color{red}novel perspective and} technique to effectively identify shifts in performance of configurable software systems. We are able to pinpoint causative commits and affected configuration options with high accuracy and tractable effort.
		\item A feasibility demonstration of our approach by implementing an adaptive learning strategy to obtain accurate estimations with acceptable measurement cost.
		\item An evaluation using both synthetic and real-world performance data from three configurable software systems. Synthetic data lets us assess our model and approach conceptually and at scale, whereas we are able to assess practicality with real-world data.
		\item A companion website\footnote{\url{https://github.com/AI-4-SE/Changepoints-Across-Variants-And-Versions} or archived at {\color{red}\url{SOFTWAR-HERITAGE}}} providing supplementary material including a reference implementation of our approach, performance measurement data, and additional visualizations.
	\end{itemize}

@@

	We illustrate the measurement effort in Figure~\ref{fig:sampling_all}, where each dot represents a pair of configuration and version.
	To reduce measurement effort, one could limit the measurements to either a few configurations (cf. Figure~\ref{fig:sampling_configurations}) or a few commits (cf. Figure~\ref{fig:sampling_versions}), but at the cost of information loss along the respective complementary dimension.
	This would leave us with either accurate performance-influence models for few versions, resulting in poor temporal granularity, or with accurate performance histories, but only for a few configurations.
	Both do not provide accurate estimations of changes in the performance influence.

@@

		\caption{Four different sampling strategies across time and configuration space.}
	\end{figure}

@@

	\subsubsection{Candidate Commits}
	For each configuration, we compute an approximation of the change point likelihood over commits (cf. Figure~\ref{fig:cpp_example}) to identify local maxima (peaks). We select such peaks under the condition that the peak change point likelihood is greater than a  threshold $t_\text{CPL}$, as defined in Equation~\ref{eq:t_cpl}. This threshold is the average change point likelihood over all commits plus a factor $N_\text{CPL}$ times its corresponding standard deviation over all commits. 
	\begin{equation}
	\begin{gathered}
	t_\text{CPL} = \frac{1}{\vert V\vert} + N_\text{CPL} \cdot \sqrt{\frac{1}{\vert V\vert} \cdot \sum_{v \in V} \Big(\frac{1}{\vert V\vert} - p'(v)\Big)^2}
		\label{eq:t_cpl}
	\end{gathered}
	\end{equation}
	The factor $N_\text{CPL}$ (default value: 3) allows us to filter peaks that do not stand out enough and might be false positives. The greater $N_\text{CPL}$ is, the stricter the filtering of peaks. That is, for each configuration in our learning set, we obtain a set of commits (configuration-specific candidate commits) that represent possible change points.
	To reduce variation among the obtained commits, we cluster the commits using kernel density estimation (KDE). The purpose of KDE in our setting is to estimate the probability mass function of whether a commit is a change point. The local maxima of this KDE, subsequently, represent our set of candidate commits.

@@

	For each parameter combination, we record the number of measurements at each iteration, the required number of iteration for termination, as well as each iteration's candidate solution. To assess the accuracy of a parameter combination's outcome (as well as intermediate iterations), we use the $\text{F}_1$ score, a combination of the metrics precision and recall. \emph{Precision} refers to the fraction of correctly identified associations (true positives) among the associations of the retrieved (candidate) solution. \emph{Recall} is the fraction of total number of the relevant associations (i.e., those we intend to find). The $\text{F}_1$ score is the harmonic mean of precision ($P$) and recall ($R$), defined as $\text{F}_1 = 2\cdot\frac{P \cdot R}{P + R}$.	In our context, we defined a correctly change point when a commits falls in a narrow 5 commit interval from the ground truth.
	To assess efficiency, we employ the required number of iterations for termination and required measurements in relation to the $\text{F}_1$ score as proxy metrics.

@@

	\subsubsection{Operationalization}
	For the experiment, we apply our approach on the performance measurements obtained in our pre-study with a few adjustments.
	First, although our study provides a broad and representative sample set of configurations, we cannot arbitrarily sample valid configurations for a couple of reasons. {\color{red}In particular, these include the commits that do not build as well as hidden configuration options and constraints. While we collected configuration options and constraints thoroughly from documentation artifacts, we can not assure that our selection is complete. We discuss this limitation further in the threats to validity.}

@@

	\section{Discussion}
	\paragraph{Efficiency} Among all experiments conducted with our synthetic setup, our approach yielded reasonable to high accuracy across most parameter settings. A subsequent analysis of parameter settings showed that our algorithm is not limited by the size of a problem space, but rather the number of measurements per iteration. Our algorithm found multiple change points across settings of up to 2,500 commits and and 2$^{64}$ configurations, resulting in a search space of about $4.6 \cdot 10^{22}$ possible measurements.
	In the context of not more than 5,000 measurements per iteration, the maximum number of measurements (after 30 iterations) added to the initial set of measurements is 150,000. For a setting with 2,500 commits, this represents around $1.4 \cdot 10^{-6}$\,\% and $3.2\cdot 10^{-16}$\,\% of possible measurements for settings with 32 and 64 options, respectively. The vast majority of cases terminated in early iterations. Hence, we are confident that our approach scales and can efficiently approximate change points for large software systems. We, we are able to reproduce this finding with real-world performance when identifying the temporal location of change points.

	\paragraph{Pinpointing Configuration Options}
	The synthetic setup has shown that our algorithm can conceptually handle both time and space. We were able to precisely identify commits for which performance changed substantially. By contrast, we were not able to fully reproduce this in a setting with real-world data regarding pinpointing change points to configuration options. Possible explanations in the case of \textsc{lrzip} include a partly inconsistent configuration interface. 
	The reason for that is not a limitation of our approach, but missing information in the real-world data (which makes the synthetic setup so important). So, we have documented that, given a precisely identified commit, we can link the affected files to configuration options with little effort. In turn, given a set of code segments of a commit, matching this with the overall code base has been extensively studied before under the umbrella of feature location techniques~\cite{hill_which_2013}.
	Inconsistencies in the documentation of configuration options is a well known and prevalent problem~\cite{rabkin_static_2011}. Our approach might be a means to complement existing feature location techniques with information about configuration-dependent performance changes.

	{\color{red}In our latter experiment, we emulated our approach on a broad but finite set of measurements rather than setting it up as an online pipeline. This does not correspond to the original layout of our algorithm, but lets us efficiently explore some limitations of this approach. We have learned that for interpreting the results of our approach, domain knowledge is advantageous as we have identified incomplete and inconsistent configuration interfaces as possible limiting factors. We acknowledge that, while our approach performs well as an automated pipeline in a controlled environment, for a real-world applications, a semi-automated setup with additional manual measurement prioritization and knowledge about configuration constraints would improve the applicability of our approach.} This is a promising further avenue of future work.

@@

	Regarding \emph{external validity}, we cannot claim that we have identified all limitations possible in a practical setting. However, we selected popular systems that are well optimized for performance and often make use of configuration options to tune performance. Moreover, by additionally performing an exhaustive synthetic study, we were able to test corner cases and assess scalability so that we believe that our results hold for many practical use cases.

>>>>

		In this work, we combine two perspectives -- variability and time -- {\color{red} into a novel perspective}. We propose an approach to identify configuration-dependent performance changes retrospectively across software variants and versions. In a nutshell, we iteratively sample pairs of configurations and versions and measure the respective performance that help us update a model of likelihoods for performance changes. Pursuing a search strategy with the goal of measuring selectively and incrementally further pairs, we increase the accuracy of identified change points related to configuration options and interactions.

@@

	Our main idea is as follows: We address the configuration complexity of this problem by selecting representative sample sets of configurations.
	Next, we uniformly sample a constant number of commits for each configuration and conduct respective performance measurements.
	Based on these measurement data, we assemble a prediction model.
	For each configuration, we estimate the likelihood of each commit being a change point, i.e. that performance of some configurations changes abruptly compared to the previous commit.
	Next, we leverage similarities in the performance histories of configurations that share common options.
	Since we sample the commits for each configuration independently, for each change point, we obtain many estimations using measurements of different commits.
	Overall, this allows us to obtain more accurate estimations of performance-changing commits with tractable effort.
	Based on a mapping from configurations to predicted change point probabilities for each configuration and commit, we derive the options responsible for each particular change point: configuration-dependent change points.
	In summary, we offer the following contributions:
	\begin{itemize}
		\item A {\color{red}novel perspective and} technique to effectively identify shifts in performance of configurable software systems. We are able to pinpoint causative commits and affected configuration options with high accuracy and tractable effort.
		\item A feasibility demonstration of our approach by implementing an adaptive learning strategy to obtain accurate estimations with acceptable measurement cost.
		\item An evaluation using both synthetic and real-world performance data from three configurable software systems. Synthetic data lets us assess our model and approach conceptually and at scale, whereas we are able to assess practicality with real-world data.
		\item A companion website\footnote{\url{https://github.com/AI-4-SE/Changepoints-Across-Variants-And-Versions} or archived at {\color{red}\url{SOFTWAR-HERITAGE}}} providing supplementary material including a reference implementation of our approach, performance measurement data, and additional visualizations.
	\end{itemize}

@@

	{\color{magenta}We illustrate the measurement effort in Figure~\ref{fig:sampling_all}, where each dot represents a pair of configuration and version.
	To reduce measurement effort, one could limit the measurements to either a few configurations (cf. Figure~\ref{fig:sampling_configurations}) or a few commits (cf. Figure~\ref{fig:sampling_versions}), but at the cost of information loss along the respective complementary dimension.
	This would leave us with either accurate performance-influence models for few versions, resulting in poor temporal granularity, or with accurate performance histories, but only for a few configurations.
	Both do not provide accurate estimations of changes in the performance influence.}

@@

		\caption{{\color{magenta}Four different sampling strategies across time and configuration space.}}
	\end{figure}

@@

	\subsubsection{Candidate Commits}
	For each configuration, we compute an approximation of the change point likelihood over commits (cf. Figure~\ref{fig:cpp_example}) to identify local maxima (peaks). We select such peaks under the condition that the peak change point likelihood is greater than a  threshold $t_\text{CPL}$, as defined in Equation~\ref{eq:t_cpl}. This threshold is the average change point likelihood over all commits plus a factor $N_\text{CPL}$ times its corresponding standard deviation over all commits. 
	\begin{equation}
	\begin{gathered}
	t_\text{CPL} = \frac{1}{\vert V\vert} + N_\text{CPL} \cdot \sqrt{\frac{1}{\vert V\vert} \cdot \sum_{v \in V} \Big(\frac{1}{\vert V\vert} - p'(v)\Big)^2}
		\label{eq:t_cpl}
	\end{gathered}
	\end{equation}
	The factor $N_\text{CPL}$ (default value: 3) allows us to filter peaks that do not stand out enough and might be false positives. The greater $N_\text{CPL}$ is, the stricter the filtering of peaks. That is, for each configuration in our learning set, we obtain a set of commits (configuration-specific candidate commits) that represent possible change points.
	To reduce variation among the obtained commits, we cluster the commits using kernel density estimation (KDE). The purpose of KDE in our setting is to estimate the probability mass function of whether a commit is a change point. The local maxima of this KDE, subsequently, represent our set of candidate commits. {\color{red}Note that if two distinct change points are almost coincident commit-wise, they can be mistakenly identified by our approach as one change point that, subsequently, can be associated with configuration options belonging to the two individual change points. }

@@

	For each parameter combination, we record the number of measurements at each iteration, the required number of iteration for termination, as well as each iteration's candidate solution. To assess the accuracy of a parameter combination's outcome (as well as intermediate iterations), we use the $\text{F}_1$ score, a combination of the metrics precision and recall. \emph{Precision} refers to the fraction of correctly identified associations (true positives) among the associations of the retrieved (candidate) solution. \emph{Recall} is the fraction of total number of the relevant associations (i.e., those we intend to find). The $\text{F}_1$ score is the harmonic mean of precision ($P$) and recall ($R$), defined as $\text{F}_1 = 2\cdot\frac{P \cdot R}{P + R}$.	In our context, we defined a correctly identified change point when a commits falls in a narrow 5 commit interval from the ground truth.
	To assess efficiency, we employ the required number of iterations for termination and required measurements in relation to the $\text{F}_1$ score as proxy metrics.

@@

	\subsubsection{Operationalization}\label{sec:rq2_operationalization}
	For the experiment, we apply our approach on the performance measurements obtained in our pre-study with a few adjustments.
	First, although our study provides a broad and representative sample set of configurations, we cannot arbitrarily sample valid configurations for a couple of reasons. {\color{red}In particular, these include the commits that do not build as well as hidden configuration options and constraints. While we collected configuration options and constraints thoroughly from documentation artifacts, we can not assure that our selection is complete. We discuss this limitation further in the threats to validity.}

@@

	\section{Discussion}
	\paragraph{Efficiency} Among all experiments conducted with our synthetic setup, our approach yielded reasonable to high accuracy across most parameter settings. A subsequent analysis of parameter settings showed that our algorithm is not limited by the size of a problem space, but rather the number of measurements per iteration. Our algorithm found multiple change points across settings of up to 2,500 commits and and 2$^{64}$ configurations, resulting in a search space of about $4.6 \cdot 10^{22}$ possible measurements.
	In the context of not more than 5,000 measurements per iteration, the maximum number of measurements (after 30 iterations) added to the initial set of measurements is 150,000. For a setting with 2,500 commits, this represents around $1.4 \cdot 10^{-6}$\,\% and $3.2\cdot 10^{-16}$\,\% of possible measurements for settings with 32 and 64 options, respectively. The vast majority of cases terminated in early iterations. Hence, we are confident that our approach scales and can efficiently approximate change points for large software systems. We, we are able to reproduce this finding with real-world performance when identifying the temporal location of change points. {\color{red}For our controlled experiment setup, we have not tuned the algorithms’ parameters to reach optimal efficiency or accuracy. However, without tuning, we demonstrated the feasibility and efficiency of our approach. Project-specific fine tuning of several parameters is a possible extension to our approach.}

	\paragraph{Pinpointing Configuration Options}
	The synthetic setup has shown that our algorithm can conceptually handle both time and space. We were able to precisely identify commits for which performance changed substantially. By contrast, we were not able to fully reproduce this in a setting with real-world data regarding pinpointing change points to configuration options. Possible explanations in the case of \textsc{lrzip} include a partly inconsistent configuration interface. 
	The reason for that is not a limitation of our approach, but missing information in the real-world data (which makes the synthetic setup so important). So, we have documented that, given a precisely identified commit, we can link the affected files to configuration options with little effort. In turn, given a set of code segments of a commit, matching this with the overall code base has been extensively studied before under the umbrella of feature location techniques~\cite{hill_which_2013}.
	Inconsistencies in the documentation of configuration options is a well known and prevalent problem~\cite{rabkin_static_2011}. Our approach might be a means to complement existing feature location techniques with information about configuration-dependent performance changes. 	
	{\color{red} In our latter experiment, instead of directly sampling new arbitrary configurations, we used a finite set of configurations as a proxy (cf.~Section~\ref{sec:rq2_operationalization}). This does not reflect the original layout of our algorithm, but lets us efficiently explore some limitations of this approach. We have learned that for interpreting the results of our approach, domain knowledge is advantageous as we have identified incomplete and inconsistent configuration interfaces as possible limiting factors. We acknowledge that, while our approach performs well as an automated pipeline in a controlled environment, for a real-world applications, a semi-automated setup with additional manual measurement prioritization and deeper knowledge of configuration constraints would improve the applicability of our approach.} This is a promising further avenue of future work.

@@

	Regarding \emph{external validity}, we cannot claim that we have identified all limitations possible in a practical setting. However, we selected popular software systems that are well optimized for performance and often make use of configuration options to tune performance.  {\color{red}The results for the experiment with these software systems do no allow for generalization, but illustrate limitations faced in a practical setting. Since these results, in conjunction with our exhaustive synthetic study, are reproducible, we were able to test corner cases and assess scalability so that we believe that our results hold for many practical use cases.}
